---
output:
  word_document: default
  html_document: default
---
title: "UBDC CCTV blog 21 04"
author: "Nick Bailey"
date: "26/04/2021"
output: html_document
---


# The easing of lockdown in Glasgow
### Nick Bailey, UBDC - `r format(Sys.Date(), "%d/%m/%Y")`

The city centres of Britain's largest urban areas have been particularly badly hit by lockdown restrictions. At the end of February, footfall in the centres of London, Birmingham, Manchester and Glasgow was down more than 80% on pre-lockdown levels, according to the Centre for Cities' 'High street recovery tracker' (https://www.centreforcities.org/data/high-streets-recovery-tracker/). Spend was down by similar amounts while the numbers of people going to work in central locations was down by even more - 96% in the case of Glasgow. The impact on retail and hospitality sectors has been enormous. 

On Monday 26th April, some of the lockdown restrictions were eased in Scotland. In particular, non-essential shops were allowed to re-open, and pubs and restaurants could offer food inside and alcohol outside. Similar changes had occurred in England two weeks earlier. 

In Glasgow, we can get a near-real-time picture of the impact of these changes thanks to a unique partnership between the Urban Big Data Centre, Glasgow City Council and the Glasgow Centre for Population Health. As explained in an earlier blog (https://www.ubdc.ac.uk/news-media/2020/july/creating-open-data-counts-of-pedestrians-and-vehicles-using-cctv-cameras/), UBDC is using spare capacity in the Council's CCTV system to produce counts of the number of pedestrians present at a range of locations across the city. Images are captured on a half-hourly cycle, and analysed within the Council's secure facilities.  All processing is done using open source tools so could be easily replicated by other authorities. If you'd like more information, please get in touch!

The resulting data are made publicly available on a daily basis through an API (see details below). In the figures below, we use these data to examine average daily counts of the number of pedestrians at different locations across the city between 10am and 4pm. Figures for the most recent week (in green) are compared with the those for the same days over the previous 16 weeks, all covered by the lockdown (in grey). 

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# load packages
pacman::p_load(httr, jsonlite, RCurl, vroom, lubridate, here, hms, tidyverse)

```

```{r data from api}

# Set date parameters for data from the API
first_date <- "2021-01-04"
last_date <- "2021-05-02"

# url with date filters
api_url <- paste("https://api.ubdc.ac.uk/cctv/counts",
                 # location,
                 first_date,
                 last_date,
                 sep="/")

# get data from api - class 'response'
cctv <- GET(url = api_url)

# get content - class 'character'
# - & replacing the parts that make it uniquely HTML
cctv_content <- content(cctv, "text") %>% 
  gsub("<html><body><p>","",.) %>% 
  gsub("</p></body></html>","",.) %>% 
  gsub("&quot;",'\"',.)

# convert to df with two cols - '.' separator in 'name'
df_cctv <- enframe(unlist(fromJSON(cctv_content)))

# split name into caseid and variable
df_cctv <- df_cctv %>% 
  separate(col = name, sep = "\\.", into = c("caseid", "variable")) 

# pivot wider, reformat, make date, remove duplicates
df_cctv <- df_cctv %>% 
  pivot_wider(caseid, names_from = "variable") %>% 
  select(-caseid) %>% 
  mutate(timestp_UTC = as_datetime(timestp_UTC)) %>% 
  mutate_at(vars(car:truck), 
            list(as.numeric)) %>% 
  mutate(date = date(timestp_UTC), 
         location_type = factor(case_when(grepl("Argyle_St", location) ~ 1, 
                                          grepl("Gallowgate", location) ~ 1,
                                          grepl("George_Sq", location) ~ 1,
                                          grepl("High_St", location) ~ 1,
                                          grepl("Hope_St", location) ~ 1,
                                          grepl("Killermont_St", location) ~ 1,
                                          grepl("Sauchiehall_St", location) ~ 1,
                                          grepl("Byres_Rd", location) ~ 2,
                                          grepl("Duke_St", location) ~ 2,
                                          grepl("Maryhill_Rd", location) ~ 2,
                                          grepl("Victoria_Rd", location) ~ 2,
                                          grepl("Park", location) ~ 3,
                                          grepl("Glasgow_Green", location) ~ 3,
                                          grepl("Broomielaw", location) ~ 4,
                                          grepl("Clyde_Walkway", location) ~ 4,
                                          grepl("Clyde_walkway", location) ~ 4,
                                          grepl("Finnieston_Bridge", location) ~ 4,
                                          grepl("Forth_Clyde_Canal", location) ~ 4,
                                          TRUE ~ 9),
                                levels = c(1,2,3,4,9), 
                                labels = c("City centre", "Other centre", 
                                           "Parks", "River/canal", "Other"))) %>% 
  distinct() 

# make list of valid locations and subset of city centre locations
counts_per_day_min <- 15 
valid_days_min <- 80

location_keep <- df_cctv %>% 
  filter(hour(timestp_UTC) >= 8 & hour(timestp_UTC) < 18) %>% 
  group_by(location, date) %>% 
  summarise(counts_per_day = n()) %>% 
  mutate(valid = (counts_per_day >= counts_per_day_min)) %>% 
  group_by(location) %>% 
  summarise(valid_days = sum(valid)) %>% 
  filter(valid_days >= valid_days_min) %>% 
  pull(location)

# # 4 selected central locations
# location_centre <- data.frame(location_keep) %>% 
#   filter(grepl("Oswald", location_keep) |
#            grepl("George_Sq", location_keep) |
#            grepl("Sauchiehall", location_keep) |
#            grepl("Waterloo", location_keep)) %>% 
#   pull()

```

## City centre
In the city centre, we can see that footfall in the week after restrictions eased was higher than all or almost all of the previous eight weeks. We are showing four locations here but a similar story is evident with all the central locations being monitored. An initial rush on the Monday was followed by a bit of a lull on the Tuesday before footfall picked up again. 

The one surprise is that footfall has not been as great as might have been expected, given the number of weeks for which the shops were closed. There is perhaps some hesitancy to return to the city centre and challenges in getting there, given we are still being encouraged to avoid public transport. 

```{r fig centre, fig.height=5, fig.width=10}

df_cctv %>% 
  filter(hour(timestp_UTC) >= 10 & hour(timestp_UTC) < 16) %>% 
  # filter(location %in% location_centre) %>% 
  filter(location %in% location_keep & 
           location_type == "City centre" &
           !grepl("Brown_St", location) &
           !grepl("cam2", location) &
           !grepl("Gordon", location)) %>% 
  group_by(location, date) %>% 
  summarise(person = mean(person), 
            N = n()) %>%
  mutate(wday = factor(wday(date, week_start = 1), 
                       labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")), 
         week = isoweek(date), 
         after_opening = (date >= date("2021-04-26"))) %>% 
  ggplot(aes(x = wday, y = person, group = week)) +
  geom_point(aes(colour = after_opening), size = 0.8) +
  geom_line(aes(colour = after_opening, size = after_opening)) +
  scale_colour_manual(values = c("lightgrey", "darkgreen")) +
  scale_size_manual(values = c(.5, 1.2)) +
  expand_limits(y=0) +
  facet_wrap(~ location, ncol = 4) +
  labs(x = "",
       y = "Daily average") +
  theme_minimal() + 
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 90)) 

```

## Other high street locations
In addition to the city centre, we have data for a number of 'high street' locations or secondary centres such as Byres Road, Duke Street, Maryhill Road and Victoria Road. Here there is much less evidence of a rise in footfall.  

```{r fig other, fig.height=3, fig.width=10}

df_cctv %>% 
  filter(hour(timestp_UTC) >= 10 & hour(timestp_UTC) < 16) %>% 
  filter(location %in% location_keep) %>% 
  filter(location_type == "Other centre") %>% 
  group_by(location, date) %>% 
  summarise(person = mean(person), 
            N = n()) %>%
  mutate(wday = factor(wday(date, week_start = 1), 
                       labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")), 
         week = isoweek(date), 
         after_opening = (date >= date("2021-04-26"))) %>% 
  ggplot(aes(x = wday, y = person, group = week)) +
  geom_point(aes(colour = after_opening), size = 0.8) +
  geom_line(aes(colour = after_opening, size = after_opening)) +
  scale_colour_manual(values = c("lightgrey", "darkgreen")) +
  scale_size_manual(values = c(.5, 1.2)) +
  expand_limits(y=0) +
  facet_wrap(~ location, ncol = 4) +
  labs(x = "",
       y = "Daily average") +
  theme_minimal() + 
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 90)) 

```

## Parks
One feature of the lockdown period has been the increased use of parks and opens spaces. With the opening up of alternatives such as shops and cafes, we might expect to see a drop in the usages of these places. That appears to be the case in Glasgow, at least for two of the largest parks - Glasgow Green and Kelvingrove. 

It should be noted that the early days of easing have been marked by a drop in temperatures compared to the previous week and that might also be a factor here. But footfall is down on almost all the previous eight weeks. At last, people have other places they can be!

```{r ig parks, fig.height=5, fig.width=10}

df_cctv %>% 
  filter(hour(timestp_UTC) >= 10 & hour(timestp_UTC) < 16) %>% 
  filter(location %in% location_keep) %>% 
  filter(location_type == "Parks" &
           !grepl("_overview", location)) %>% 
  group_by(location, date) %>% 
  summarise(person = mean(person), 
            N = n()) %>%
  mutate(wday = factor(wday(date, week_start = 1), 
                       labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")), 
         week = isoweek(date), 
         after_opening = (date >= date("2021-04-26"))) %>% 
  ggplot(aes(x = wday, y = person, group = week)) +
  geom_point(aes(colour = after_opening), size = 0.8) +
  geom_line(aes(colour = after_opening, size = after_opening)) +
  scale_colour_manual(values = c("lightgrey", "darkgreen")) +
  scale_size_manual(values = c(.5, 1.2)) +
  expand_limits(y=0) +
  facet_wrap(~ location, ncol = 4) +
  labs(x = "",
       y = "Daily average") +
  theme_minimal() + 
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 90)) 

```


## The riverfront
Glasgow's riverfront has also been a popular destination over lockdown for walking, cycling or just hanging out. As with the parks, footfall has been lower this week overall. 

```{r fig river, fig.height=3, fig.width=10}

df_cctv %>% 
  filter(hour(timestp_UTC) >= 10 & hour(timestp_UTC) < 16) %>% 
  filter(location %in% location_keep) %>% 
  filter(location_type == "River/canal" &
           !grepl("Watt", location)) %>% 
  group_by(location, date) %>% 
  summarise(person = mean(person), 
            N = n()) %>%
  mutate(wday = factor(wday(date, week_start = 1), 
                       labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")), 
         week = isoweek(date), 
         after_opening = (date >= date("2021-04-26"))) %>% 
  ggplot(aes(x = wday, y = person, group = week)) +
  geom_point(aes(colour = after_opening), size = 0.8) +
  geom_line(aes(colour = after_opening, size = after_opening)) +
  scale_colour_manual(values = c("lightgrey", "darkgreen")) +
  scale_size_manual(values = c(.5, 1.2)) +
  expand_limits(y=0) +
  facet_wrap(~ location, ncol = 4) +
  labs(x = "",
       y = "Daily average") +
  theme_minimal() + 
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 90)) 

```

### Further information
More information on our work using CCTV to assess daily activity on city streets is available here: https://www.ubdc.ac.uk/research/research-projects/urban-sensing-analytics/using-spare-cctv-capacity-to-monitor-activity-on-city-streets/. Current developments include enhancements to the capture of vehicles and cycling, as well as the development of video analysis to let us capture behaviours in more detail - for example, whether people spend time in certain places or just pass through. 

The data from the CCTV analysis are available here as open data: https://api.ubdc.ac.uk/cctv/. They are made available under an Open Government Licence. The attribution is: 

"Glasgow CCTV Automated Object Detection Counts, Glasgow City Council / Urban Big Data Centre at the University of Glasgow, 2020, copyright © Glasgow City Council 2020"

The code to re-produce this analysis is available here: https://github.com/nick-bailey/UBDC-CCTV-data. 

### Project Team
The data examined here are the result of a project involving a team of people, including: 

* UBDC: Dr Mark Livingston (lead); Luis Serra; Dr David McArthur; Dr Andrew McHugh; Maralbek Zeinullin
* Glasgow City Council: Kimberley Hose; Keith Scott; Kalim Uddin
* Glasgow Centre for Population Health: Bruce Whyte