---
title: "UBDC_CCTV_GALLANT_v1"
author: "Nick Bailey"
date: "14/08/2023"
output:
  html_document:
    code_folding: hide
---


# UBDC CCTV data - GALLANT v1 - `r Sys.Date()`

Data extraction and analysis for GALLANT case study neighbourhoods, August 2022 to July 2023, covering 14 cameras. 


```{r setup}

#knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)


# load packages
pacman::p_load(zoo, httr, jsonlite, curl, gt, feather, here, 
               lubridate, ggrepel, viridis, tidyverse, readxl)

```


## Info on cameras
Retrieve information on the camera locations and identities from spreadsheet. 

```{r cameras cs}

# cameras in GALLANT case study nhds - spreadsheet from Emma L.
df_cs_cameras <- read_xlsx(here('GALLANT', 'df_cameras GALLANT WP4.xlsx'), 
                           sheet = 'df_cameras') %>% 
  rename(cs_nhd = `GALLANT nhd`) %>% 
  filter(!is.na(cs_nhd)) %>% 
  select(camera, camera_name, cs_nhd) %>% 
  mutate(camera_name = paste0(camera_name, " (", camera, ")")) 

# extract camera numbers
cs_cameras = unique(df_cs_cameras$camera)

```



## Load data
Load data from the API for the relevant cameras. 

```{r data cs}

# set dates - last 12 month
first_date <- "2022-08-01"
last_date <- "2023-07-31"


df_cctv <- data.frame()

for (i in 1:length(cs_cameras)) {

  # base url with date filters
  # - reformat dates for query
  # - page_size set to max (1000)
  api_url <- paste0("https://glasgow-cctv.ubdc.ac.uk/api/records/",
                    "?date_after=", gsub("-", "", first_date),
                    "&date_before=", gsub("-", "", last_date),
                    "&camera_id=", as.character(cs_cameras[i]),
                    "&format=json",
                    "&page_size=1000")

  # get first page of data from api
  cctv <- GET(url = api_url)

  # get content
  cctv_content <- content(cctv, "text") %>%
    gsub("<html><body><p>","",.) %>%
    gsub("</p></body></html>","",.) %>%
    gsub("&quot;",'\"',.)

  # extract records from json
  temp <- enframe(unlist(fromJSON(cctv_content)))

  # make number of pages - n. records/1000, rounded up
  n_pages <- temp %>%
    filter(name == "count") %>%
    mutate(value = trunc(as.numeric(value)/1000) + 1) %>%
    pull(value)


  # list of pages
  ls_pages <- as.list(paste0(api_url, "&page=", 1:n_pages))

  # adapted from:
  # - https://www.roelpeters.be/three-ways-to-make-asynchronous-get-requests-in-r/
  # - https://cran.r-project.org/web/packages/curl/vignettes/intro.html
  pl <- new_pool()
  bodies <- list()

  # fn for api content
  cb <- function(x) {
    bodies <<- append(bodies,
                      rawToChar(x$content)
                      )
  }

  # create requests
  lapply(ls_pages, function(x) { curl_fetch_multi(x, done = cb, pool = pl) })

  # perform requests
  out <- multi_run(pool = pl)

  # # check returns
  # print(out)

  # convert to list of dfs
  bodies <- lapply(bodies, function(x) {enframe(unlist(fromJSON(x)))})

  # convert to single df
  temp1 <- bind_rows(bodies, .id = "column_label") %>%
    filter(name != "count" &                # remove lines which are not records
             name != "next" &
             name != "previous") %>%
    mutate(name = gsub("results.", "", name)) %>%   # remove leading text
    separate(name,                                  # split text from numeric
           into = c("var", "case"),
           sep = "(?<=[A-Za-z])(?=[0-9])"
           ) %>%
    mutate(case = (as.numeric(column_label)-1) * 1000 + as.numeric(case)) %>% 
    mutate(case = case * 100 + i) %>%   # ensure no duplicates between cameras
    select(-column_label)

  df_cctv <- df_cctv %>%
    rbind(temp1)

}


# further cleaning
df_cctv <- df_cctv %>% 
  pivot_wider(names_from = "var", values_from = "value") %>% 
  mutate_at(vars(cars:buses), 
            list(as.numeric)) %>% 
  mutate(timestamp = as_datetime(timestamp),
         hour = hour(timestamp),
         date = date(timestamp)) %>% 
  mutate(time = factor(case_when(hour >=  6 & hour < 10 ~ 1,
                                 hour >= 10 & hour < 16 ~ 2,
                                 hour >= 16 & hour < 18 ~ 3,
                                 hour >= 18 & hour < 22 ~ 4,
                                 hour >= 22 | hour < 06 ~ 5),
                       levels = 1:5,
                       labels = c("AM peak (6am-10am)",
                                  "Daytime (10am-4pm)",
                                  "PM peak (4pm-6pm)",
                                  "Evening (6pm-10pm)",
                                  "Night  (10pm-6am)"))) %>%
  mutate(date2 = case_when(hour < 04 ~ (date - 1),  # midnight-3am in prev day
                          TRUE ~ date),
         day = wday(timestamp, week_start = 1), 
         weekend = (day >= 6), 
         month = month(timestamp)) %>% 
  mutate(camera = as.numeric(camera))


# add names and nhd
df_cctv <- df_cctv %>% 
  left_join(df_cs_cameras, by = 'camera')


# for figs
latest_date <- paste0(day(max(df_cctv$date)), "/", month(max(df_cctv$date)), 
                         "/", year(max(df_cctv$date)))


# col names for counts
count_cols <- c("cars", "persons", "bicycles", "trucks", "motorcycles", "buses")


# drop cols, reorder and tidy
df_cctv <- df_cctv %>% 
  select(-case, -id, -camera_ok, -model_name) %>% 
  select(camera, cs_nhd, camera_name, 
         timestamp, date, hour, day, 
         everything()) %>% 
  pivot_longer(cols = all_of(count_cols), 
               names_to = 'object', 
               values_to = 'count')


# tidy up
rm(cctv)
rm(bodies)
rm(out)
rm(ls_pages)

```

## Hourly dataframe
Counts are now hourly averages. Filling in gaps through rolling averages for hours either side, or weeks either side. 

```{r hourly, warning=FALSE}

# hourly-level dataset from original
df_cctv_hourly <- df_cctv %>%
  group_by(object, camera, date, hour) %>%
  summarise(count = mean(count, na.rm = TRUE),    # counts are now hourly averages
            N = n(), 
            present = (n() > 0)) 


# expand grid 
df_cctv_hourly <- expand_grid(object = unique(df_cctv_hourly$object), 
                            camera = unique(df_cctv_hourly$camera), 
                            date = seq(min(df_cctv_hourly$date), max(df_cctv_hourly$date), by = "1 day"), 
                            hour = unique(df_cctv_hourly$hour)) %>% 
  left_join(df_cctv_hourly, by = c('object', 'camera', 'date', 'hour')) %>% 
  mutate(day = wday(date, week_start = 1))


# missing N/present set to zero
df_cctv_hourly$N[is.na(df_cctv_hourly$N)] <- 0
df_cctv_hourly$present[is.na(df_cctv_hourly$present)] <- 0


# first and last date for each camera
df_first_last_date <- df_cctv %>%
  mutate(date = date(timestamp)) %>%                # put date/day back to original
  group_by(camera) %>% 
  summarise(date_first = min(date), 
            date_last = max(date))


# cut grid to first/last date for each camera
df_cctv_hourly <- df_cctv_hourly %>% 
  left_join(df_first_last_date, by = 'camera') %>% 
  filter(date >= date_first & date <= date_last)


# make rolling measures - hours either side
df_cctv_hourly <- df_cctv_hourly %>% 
  arrange(object, camera, date, hour) %>% 
  group_by(object, camera) %>% 
  mutate(count_hr3 = rollmean(count, k=3, fill = "fill", na.rm = TRUE), 
         count_hr5 = rollmean(count, k=5, fill = "fill", na.rm = TRUE), 
         present_hr3 = (rollmean(present, k=3, fill = "fill") > 0),
         present_hr5 = (rollmean(present, k=5, fill = "fill") > 0))


# make rolling measures - weeks either side
df_cctv_hourly <- df_cctv_hourly %>% 
  mutate(day = wday(date, week_start = 1)) %>% 
  arrange(object, camera, day, hour, date) %>% 
  mutate(count_wk3     = rollmean(count, k=3, fill = "fill", na.rm = TRUE), 
         count_hr3_wk3 = rollmean(count_hr3, k=3, fill = "fill", na.rm = TRUE), 
         count_hr5_wk5 = rollmean(count_hr5, k=5, fill = "fill", na.rm = TRUE),
         present_wk3     = (rollmean(present, k=3, fill = "fill") > 0), 
         present_hr3_wk3 = (rollmean(present_hr3, k=3, fill = "fill") > 0),
         present_hr5_wk5 = (rollmean(present_hr5, k=5, fill = "fill") > 0)) %>% 
  arrange(object, camera, date, hour)


# make combined measure using 'best available' and associated quality measure
df_cctv_hourly <- df_cctv_hourly %>% 
  mutate(count_combined = case_when(!is.na(count)         ~ count, 
                                    !is.na(count_hr3)     ~ count_hr3,
                                    !is.na(count_hr5)     ~ count_hr5,
                                    !is.na(count_wk3)     ~ count_wk3,
                                    !is.na(count_hr3_wk3) ~ count_hr3_wk3,
                                    !is.na(count_hr5_wk5) ~ count_hr5_wk5), 
         quality_combined  = case_when(!is.na(count)      ~ 6, 
                                    !is.na(count_hr3)     ~ 5,
                                    !is.na(count_hr5)     ~ 4,
                                    !is.na(count_wk3)     ~ 3,
                                    !is.na(count_hr3_wk3) ~ 2,
                                    !is.na(count_hr5_wk5) ~ 1, 
                                    TRUE ~ 0))

# # check on cameras
# table(df_cctv_hourly$camera)

```

## Daily data frame
Count is now daily total

```{r daily}

df_cctv_daily <- df_cctv_hourly %>% 
  group_by(camera, object, date) %>% 
  summarise(count = mean(count_combined, na.rm = TRUE) * 24,   # in case there are any missing values still
            N = sum(N, na.rm = TRUE),
            quality = mean(quality_combined, na.rm = TRUE)) %>% 
  left_join(df_cs_cameras, by = 'camera')  %>% 
  mutate(day = wday(date, week_start = 1), 
         month = month(date), 
         year = year(date),
         weekend = (day >= 6)) %>% 
  select(cs_nhd, camera, camera_name, object, date, year, month, day, weekend, count, N, quality, everything()) %>% 
  arrange(cs_nhd, camera, camera_name, object, date)

```


## Monthly data frame
Count is now daily average.

```{r monthly}

df_cctv_monthly <- df_cctv_daily %>% 
  group_by(cs_nhd, camera, camera_name, object, weekend, year, month) %>% 
  summarise(yrmon = as.yearmon(first(date)),
            count = mean(count, na.rm = TRUE),
            N_days = n(),
            quality = mean(quality, na.rm = TRUE)) %>% 
  arrange(cs_nhd, camera, camera_name, object, weekend, year, month)

# # check on cameras
# table(df_cctv_monthly$camera)

temp <- df_cctv_monthly %>% 
  pivot_wider(names_from = 'object', values_from = 'count')

write_csv(temp, here("GALLANT", "GALLANT case study monthly counts.csv"))

```


## Coverage of cameras

Shows the number of images on each day which go into the daily average. 


```{r counts 1, fig.height=15, fig.width=10}

# average quality measure - denominator is all days in range
df_cctv_daily %>% 
  filter(object == 'cars') %>% 
  ggplot(aes(x = date, y = N)) +
  geom_line() +
  scale_x_date(breaks = 'month') +
  theme_minimal() +
  theme(panel.grid.minor.x = element_blank() ) +
  facet_wrap(~ camera_name, ncol = 2) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = 'No. of images per day')


```

Shows the number of days in each month for which camera has data. 


```{r counts 2, fig.height=10, fig.width=10}

# average quality measure - denominator is all days in range
df_cctv_monthly %>% 
  filter(object == 'cars') %>% 
  ggplot(aes(x = yrmon, y = N_days)) +
  geom_line(aes(colour = weekend), size = 1.2) +
  scale_x_yearmon(n = 15) +
  theme_minimal() +
  facet_wrap(~ camera_name, ncol = 3) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = 'No. of days with data each month')


```

## Results


Weekday totals

```{r weekday 1, fig.height=6, fig.width=10}

df_cctv_monthly %>% 
  filter(weekend == 0) %>% 
  ggplot(aes(x = yrmon, y = count)) +
  geom_line(aes(colour = camera_name), size = 1.2) + 
  expand_limits(y = 0) +
  scale_x_yearmon(n = 15) +
  facet_wrap(~ object, scales = 'free') +
  theme_minimal() +
  theme(panel.grid.minor.x = element_blank() ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Weekday daily average", 
       x = 'Month',
       y = 'Daily average count',
       caption = "Source: UBDC CCTV data")

```

```{r weekend 1, fig.height=6, fig.width=10}

df_cctv_monthly %>% 
  filter(weekend == 1) %>% 
  ggplot(aes(x = yrmon, y = count)) +
  geom_line(aes(colour = camera_name), size = 1.2) + 
  expand_limits(y = 0) +
  scale_x_yearmon(n = 15) +
  facet_wrap(~ object, scales = 'free') +
  theme_minimal() +
  theme(panel.grid.minor.x = element_blank() ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Weekend daily average", 
       x = 'Month',
       y = 'Daily average count',
       caption = "Source: UBDC CCTV data")

```


```{r weekday 2, fig.height=10, fig.width=10}

df_cctv_monthly %>% 
  filter(weekend == 0) %>% 
  ggplot(aes(x = yrmon, y = count)) +
  geom_line(aes(colour = object), size = 1.2) + 
  expand_limits(y = 0) +
  scale_x_yearmon(n = 15) +
  facet_wrap(~ camera_name, ncol = 3, scales = 'free_y') +
  theme_minimal() +
  theme(panel.grid.minor.x = element_blank() ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Weekday daily average", 
       x = 'Month',
       y = 'Daily average count',
       caption = "Source: UBDC CCTV data")

```

```{r weekend 2, fig.height=10, fig.width=10}

df_cctv_monthly %>% 
  filter(weekend == 1) %>% 
  ggplot(aes(x = yrmon, y = count)) +
  geom_line(aes(colour = object), size = 1.2) + 
  expand_limits(y = 0) +
  scale_x_yearmon(n = 15) +
  facet_wrap(~ camera_name, ncol = 3, scales = 'free_y') +
  theme_minimal() +
  theme(panel.grid.minor.x = element_blank() ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Weekend daily average", 
       x = 'Month',
       y = 'Daily average count',
       caption = "Source: UBDC CCTV data")

```